\maketitle{}

\begin{abstract}
  Futhark is a purely functional programming language, designed to be compiled
  to efficient GPU code.
  The GPU code has both C and Python interfaces, but does not interoperate with
  other mainstream languages such as \csharp{}, \fsharp{} or Java.

  We first describe and implement a \csharp{} code generator that allows GPU kernels
  written in Futhark to be compiled and used as external libraries in \csharp{}-
  and \fsharp{} programs.

  We then describe and implement the language \fshark{}, which lets developers
  write and prototype their own GPU code entirely as declarative \fsharp{} code.
  The language comes with a compiler of its own, which compiles \fshark{}
  programs to Futhark code.

  We argue that the \fshark{}-to-Futhark translations are correct by comparing
  the results of \fsharp{}-evaluated \fshark{} code with their Futhark-evaluated
  counterparts, by running a comprehensive suite of 128 unit tests.

  We show that Futhark can be compiled to \csharp{} GPU code, resulting in
  comparable performance with the \clang{} OpenCL backends ($0-3\%$ for complex benchmarks),
  by comparing runtimes across over 20 Futhark benchmark programs ported from other array
  programming languages.

  Finally, we show that complex GPU benchmarks can be easily written in
  \fshark{} in a way that preserves the idiomatic \fsharp{} style.
  The GPU benchmarks are compiled to GPU code and executed in an \fsharp{}
  context, where they are shown to perform comparably with benchmarks handwritten in
  Futhark.
\end{abstract}
 
\tableofcontents

\chapter*{Preface}
This thesis is submitted in fulfillment of a 30 ECTS master's thesis in Computer
Science (Datalogi) at the University of Copenhagen, for Mikkel Storgaard Knudsen.

(Front page picture: The humble hedgehog is a peaceful omnivore firmly placed in
the lower-middle part of the food chain. However, sneaky disguises occasionally allows it
to partake in feasts that are otherwise reserved for apex predators.)

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../thesis"
%%% End: